Convergent Cellular Adaptation to Freeze-Thaw Stress via a Quiescence-like State in Yeast

Version 1

Abstract

Extreme stress, such as freeze-thaw, poses a severe challenge to many organisms, but the mechanisms underlying their adaptation to survive such stress remain elusive. Here, we show that Saccharomyces cerevisiae can rapidly evolve freeze-thaw tolerance through a physiological state transition, with survival increasing nearly two orders of magnitude from ≈2% to ≈70% in about 25 cycles of stress exposure. Evolved yeast cells exhibit a quiescence-like state, characterized by altered cellular physiology: increased intracellular trehalose accumulation, reduced membrane damage, cytoplasmic stiffening and an exit from a proliferative cycle. This mechano-chemically reinforced survival strategy emerges across independent evolutionary lines despite distinct genetic backgrounds, suggesting a convergent mechanism of adaptation. By integrating experimental evolution, biophysical measurements, genomic analysis, and a quantitative model that captures the adaptation dynamics, we reveal that stress tolerance can arise via a potentially generalizable, physiologically mediated adaptation strategy. These findings provide new insights into microbial survival under extreme conditions and suggest broader implications for cellular stress responses beyond yeast.

Introduction

Cells across many domains of life survive extreme environmental perturbations, including complete desiccation and freezing, which can cause severe cellular damage. Cellular damage occurs via many mechanisms such as protein denaturation [1, 2], membrane rupture [3, 4], and oxidative stress [5]. To survive such conditions, organisms have evolved diverse adaptive mechanisms [6, 7]. Classical evolutionary models primarily emphasize genetic mutations as the basis for the evolution of such adaptations [8–10]. However, recent evidence suggests that cells can also adapt via phenotypic plasticity and physiological state transitions, allowing rapid, non-genetic survival advantages [11–16].

One widely recognized mechanism of stress resistance is the accumulation of protective molecules such as trehalose, which stabilizes proteins and membranes during dehydration and freezing [14, 17, 18]. Additionally, molecules such as trehalose have also been associated with transition to quiescence, a reversible state of metabolic suppression, to endure unfavorable conditions, in many microbes and multicellular organisms. Among many other characteristics, quiescence is characterized by increased cytoplasmic viscosity, enhanced resistance to oxidative damage, and improved long-term viability [19–22]. In microbes, quiescence has been linked to persistence under starvation and antimicrobial stress, while in multicellular organisms, it plays a role in stem cell maintenance and tumor dormancy [23–25].

Here, we investigate whether repeated exposure to extreme freeze-thaw stress selects for a quiescence-like adapted state in Saccharomyces cerevisiae. Using experimental evolution, we subjected yeast populations to repeated cycles of rapid freezing and thawing, a severe mechano-chemical perturbation that initially results in very low survival (<2%) [26, 27]. Strikingly, within ∼25 selection cycles, survival increased nearly 40-fold. The adapted yeast cells exhibit an increased prevalence of a quiescence-like state, characterized by altered cellular physiology: increased intracellular trehalose accumulation, reduced membrane damage, cytoplasmic stiffening and an exit from a proliferative cycle. Whole-genome sequencing revealed that the adaptation was not driven by unique genetic pathways but rather by multiple, distinct genetic backgrounds, all converging on the same mechano-chemically reinforced survival phenotype.

Our findings suggest that a quiescence-like adaptation represents a generalizable stress tolerance strategy. While our results strongly suggest a dominant role for non-genetic physiological adaptation, it is possible that genetic and regulatory factors may contribute to this response. These results raise intriguing questions about whether similar physiological state transitions contribute to microbial persistence in fluctuating environments [28] and whether such mechanisms can be harnessed for industrial and medical applications, including bio-preservation and cryo-tolerance. By demonstrating that cells can evolve resilience via a physiologically mediated mechanism that operates alongside genetic factors, our work expands the understanding of adaptive evolution and stress biology beyond classical mutationdriven models [29].

Results

Saccharomyces cerevisiae adapts to freeze-thaw stress

We first establish an experimental protocol in order to perform the selection (Methods; Figure 1A). It is known that the survival of Saccharomyces cerevisiae upon freeze-thaw depends on the rate of freezing and thawing. Survival has been found to be minimal when it undergoes a fast freeze (> 10 K/min) to −196°C [?]. We chose this regime to work in to possibly achieve the most significant increase in survival after adaptation. Before proceeding, we determined the effect of duration at −196°C on survival by exposing the WT (CEN.PK a/ α) stationary phase cell cultures to a fast freeze (> 10K/min) to −196°C and allowed them to remain at −196°C for varying lengths of time before thawing. Freezing was done by immersing 500 µl of stationary phase culture in 1.5 mL tubes in liquid nitrogen (−196°C) and the cultures were then thawed by immersing them in 25°C water for 15 minutes. We note that the survival fraction of yeast does not vary with duration at −196°C (Supplementary Information). Therefore, for experimental convenience, we used a freeze duration of 15 minutes and selected for freeze-thaw tolerance at high cooling rates (> 10 K/min).

Experimental evolution can be used to increase survival of yeast under freeze-thaw stress.

A Experimental design of selection for freeze-thaw tolerance. B Survival fraction of cells after freeze-thaw over the course of evolution. Inset: Number of viable cells after every freeze-thaw-growth cycle. Purple - evolving population. Pink - control population. C Basal intracellular concentrations of trehalose relative to the control measured over the course of evolution.

The cells that survived the freeze-thaw were grown in fresh medium until stationary phase (24 hours, ≈4 generations) and this cycle repeats. The survival fraction was then measured by colony forming units (CFUs). This cycle of freezethaw followed by growth was repeated until the survival fraction stopped increasing. For controls, the same protocol was followed without the freeze-thaw procedure. To start the experiments, a single yeast colony was picked from an agar plate and grown as liquid culture to stationary phase. This starter culture was divided into six parts — two for controls and four biological replicates for selection.

Under this selection protocol, the survival fraction increased from ≈2% up to an average of ≈70%, with one of the replicates as high as 85%, in only a span of ≈25 freeze-thaw cycles. This shows that the yeast cells adapt quite rapidly to freeze-thaw in this regime of very low initial survival rates (i.e. ≈2%) and the dynamics are remarkably similar across independent evolution lines (Figure 1B). Next, we characterized the differences in the evolved strain with respect to the WT.

Freeze-thaw tolerant populations of cells have higher basal intracellular levels of trehalose

Trehalose is a sugar containing two α-glucose units with a 1,1-glycosidic bond between them. Several studies have found a strong correlation between intracellular trehalose concentration and survival after freeze-thaw [30, 31]. Chen et al. showed that trehalose was sufficient to confer freeze-thaw tolerance when cooled slowly to −80°C [32]. These studies either measure the amount of trehalose in existing freeze-thaw tolerant strains or the survival in trehalose accumulated mutant strains. Motivated by these studies, we measured the amounts of intracellular trehalose at the stationary phase to calculate the basal levels of trehalose present in cells, over the course of adaptation to freeze-thaw. Cell colonies were collected and weighed from plates used to measure survival using CFUs 48 hours after thawing. The concentration of trehalose was then measured using a previously described protocol [33] (Methods). We found that the amount of basal intracellular trehalose systematically increased over the course of adaptation to freeze-thaw, exhibiting similar dynamics to those of the cellular survival levels (Figure 1C). Basal levels of trehalose increased nearly 3 fold over the course of the selection.

Previous studies suggest that trehalose might play a role in reducing membrane damage [?], which has been implicated as the primary cause of death after freeze-thaw [34–36]. Therefore, we next turned our attention to investigating if the freeze-thaw tolerant cells have a membrane-damage related adaptation.

Freeze-thaw tolerant cells undergo reduced membrane damage

The presence or absence of membrane damage was measured by staining the cells with two dyes — 5-carboxyfluorescein diacetate (5-CFDA) and Propidium Iodide (PI) — and measuring the intensities of fluorescence of both dyes using flow cytometry (Methods). The fluorescent form of CFDA leaks out of a cell quickly if it is membrane damaged. PI permeates cells with damaged membranes. Taken together, these dyes allow a clear distinction between membranedamaged cells and cells with intact membranes, showing up as two distinct clusters in the data (Figure 2A,B). Each point on these plots represents a single cell. We compute a single metric that combines this information to estimate the total membrane damage (Methods), which clearly indicates that a smaller fraction of freeze-thaw tolerant cells undergoes membrane damage after freeze-thaw (Figure 2C). Further, we sought to understand the effect of membrane damage on cellular survival. We used fluorescence-activated cell sorting (FACS) to sort the cells into “membranedamaged” and “membrane-intact” fractions. We then measured the number of colony forming units (CFUs) within each sorted fraction to quantify their viability (detailed description in Supplementary Information). We found that the survival fraction of freeze-thaw tolerant cells within the membrane-damaged sub-population exceeded the wild type by about two orders of magnitude (Figure 2D). Similarly, the viability of freeze-thaw tolerant cells with intact membranes was an order of magnitude higher than the corresponding wild type population.

Freeze-thaw tolerant cells undergo reduced membrane damage and survive membrane damage.

A Flow cytometry measurements on Wild Type cells. Each point represents a single cell and the plot contains data for 10000 cells. The x-axis is the fluorescence intensity in the 5-CFDA channel. The y-axis is the fluorescence intensity in the propidium iodide channel. The contour lines represent changes in probability density of cells. The cluster on the top left comprises membrane damaged cells and the cluster at the bottom right comprises cells with intact membranes. B Flow cytometry measurements on freeze-thaw tolerant cells. Each point represents a single cell and the plot contains data for 10000 cells. The x-axis is the fluorescence intensity in the 5-CFDA channel. The y-axis is the fluorescence intensity in the propidium iodide channel. The contour lines represent changes in probability density of cells. The cluster on the top left comprises membrane damaged cells and the cluster at the bottom right comprises cells with intact membranes. C The 2-dimensional data from A and B are used to compute a 1-dimensional distribution of “membrane damage” in cells. D Viability of cells measured after sorting the two clusters identified from the data in A and B.

These data indicate that: (i) a significant fraction of cellular death following freeze-thaw is unrelated to membrane damage and (ii) the evolved cells have mechanisms that can aid in survival following membrane damage. This suggests cellular mechanisms and attributes beyond just the membrane protective roles of trehalose.

Increased intracellular trehalose is a defining characteristic of cellular quiescence — a temporary non-proliferating state — during which cells are tolerant to a range of stress conditions [37]. Therefore, we next investigated if the freeze-thaw tolerant cells exhibit traits similar to those of quiescent cells [38]. We started by probing the morphological and physical traits of freeze-thaw tolerant cells in the stationary phase – the point in their population growth stage at which the freeze-thaw stress was imposed.

Freeze-thaw tolerant cells are smaller, denser, and exhibit reduced budding

To investigate if there could be any morphological changes associated with the increased survival of cells, we imaged cells of the wild-type yeast and the freeze-thaw tolerant yeast at stationary phase to measure their size (Figure 3A). We observed a two-fold reduction in cell area (Figure 3B), which corresponds to an estimated six-fold volume reduction in freeze-tolerant cells. A decrease in cell volume could be due to a decrease in its mass and/or increase in cell density. By measuring the single cell mass density of cells at stationary phase using a density gradient column, we found that the freeze-thaw tolerant cells had a significantly higher cell density (Figure 3C).

Cells of the population adapted to survive freeze and thaw are smaller, denser and exhibit reduced budding.

A Brightfield image of cells. Scale bar - 10µm. B Projected area of cells measured from images as shown in A. Sample size: Control = 1107, Evolved = 4281, P = 1.387 ∗ 102. C Single cell mass density measured using a percoll gradient column. P = 1.7867 ∗ 107. D Fraction of cells budding in stationary phase measured from images as shown in A. Sample size: Control = 309, Evolved = 422, P = 1.2079 ∗ 106. E Distribution of normalized fluorescence intensities of DNA in cells stained with SYTOX Green measured using flow cytometry. Sample size: Control = 199682, Evolved = 399113. F Design of the microfluidic channel through which cells are flowed to measure deformation with an image of cells leaving the microfluidic channel. Scale bar - 15µm. G Eccentricity and area of cells measured in the narrow channel. Flow rate: Control - 5.5cm/s, Evolved - 5.6cm/s. H Fraction of intact cells over time measured by turbidity of cells placed in water after Zymolyase treatment. I Image of Genetically Encoded Multimers (GEMs) in the cytoplasm of cells. Scale bar - 5µm. J Mean squared displacement (MSD) of GEMs tracked in the cytoplasm of cells, plotted on logarithmic axes. Inset: MSD plotted on linear axes. K Distribution of the exponent of time, α.

From the cellular morphologies, we also found a reduction in the incidence of budding in the evolved cell population (Figure 3D), suggesting that they stopped doubling before reaching stationary phase and could have exited the proliferating phase of the cell cycle. This is also corroborated by a measurement of total amount of cellular DNA — the evolved cell population has reduced fraction of dual DNA copies (Figure 3E), suggestive of quiescence.

The increased cellular mass density suggests that there could be an associated increase in molecular crowding in the cytoplasm, thereby reducing cytoplasmic mobility and increasing cellular stiffness. Since we also observed reduced membrane damage, we investigated if the freeze-thaw tolerant cells are more mechanically resilient.

Freeze-thaw tolerant cells are mechanically stiffer

Yeast cells are naturally ellipsoid due to the presence of a stiff cell wall. Exposure to Zymolyase, a cell-wall digesting enzyme, leads, over time, to the formation “spheroplasts”, spherical cells bound by a lipid membrane. We used these Zymolyase treated cells to quantify their stiffness, by employing deformability cytometry a method previously described by Otto et al. [39]. Briefly, the cytometry is performed by suspending the Zymolyase treated cells in a viscous buffer and flowing them through a narrow microfluidic channel. The resultant shear causes cellular deformation, which is measured using high-speed imaging (Figure 3F). The deformation of cells due to shear is quantified by their eccentricity in relation to their size. For a given cell size, lower cell stiffness results in a higher deformation and, therefore, higher eccentricity. From such a measurement, we found two clusters of freeze-thaw tolerant cells with a majority of cells minimally deformed, compared to the freeze-thaw sensitive (WT) cells (Figure 3G). The presence of two clusters indicates increased cellular stiffness even after the long Zymolyase treatment. Munder et al. found that cells that had increased cytoplasmic stiffness did not relax into spheres after treatment with Zymolyase [19]. Hence, the relaxation of freeze-thaw tolerant cells from ellipsoidal to spherical upon treatment with Zymolyase could be affected either by their cytoplasmic stiffness or their cell wall resilience due to a varied cell wall composition that is not affected by Zymolyase to the same extent. Therefore, we further investigated the role of these two factors – cell wall resilience and cytoplasmic stiffness – in increased cellular stiffness.

In yeast, the cell wall maintains the structure of the cell. When the cell wall is removed and cells are placed in a hypo-osmotic environment, cells lyse as they cannot regulate the amount of water permeating the cell. Therefore, to investigate the presence of the cell wall after digestion with Zymolyase, we exposed cells to Zymolyase for varying lengths of time in an appropriate buffer followed by a hypo-osmotic environment i.e. pure water (Supplementary Information). Cells lacking cell-wall lyse in water. So, we measured the turbidity of the solution after placing the cells in water as a measure of the fraction of intact cells. A significant fraction of the freeze-thaw tolerant cells do not lyse even after exposure to Zymolyase for long periods of time, compared to the duration taken for most of the freeze-thaw sensitive cells to lyse (Figure 3H). This suggests three possibilities - (a) the freeze-thaw tolerant cells have an altered cell wall composition that is not easily digested by Zymolyase, (b) the cells have an altered membrane structure (as seen above) that does not allow water to enter the cell even after cell wall digestion, or (c) an increased cytoplasmic stiffness that does not allow cells to relax after cell wall digestion and hence the cellular components do not spread into solution enough to change the turbidity of the solution.

These results suggest that cells of the adapted population have increased mechanical (structural) resilience associated with modified cell wall or cytoplasmic properties. Next, we specifically investigated the cytoplasmic rheology to understand the nature of the increased cellular stiffness.

Freeze-thaw tolerant cells exhibit reduced cytoplasmic mobility

To study the cytoplasmic mobility, we tracked the intracellular motion of engineered fluorescent particles (41 nm diameter) (Figure 3I). Briefly, to produce these fluorescent particles, we transformed the wild type CEN.PK strain to express genetically encoded multimers (GEMs) using a plasmid developed by Delarue et al. [40]. Delarue et al. replaced the INO4 promoter with the monomeric fluorophore sequence fused with the INO4 promoter and selected for the integrants using uracil auxotrophy. We inserted the monomeric fluorophore fused with the INO4 promoter at the HO locus in a prototroph which does not affect growth characteristics significantly in a diploid strain [41]. The plasmid also contained a KanMX marker for selecting the integrants using resistance to G418 (full details in the Supplementary Information). We imaged the GEMs expressed in the cell using total internal reflection fluorescence (TIRF) microscopy, used image analysis to track the particles and analysed the trajectory dynamics using custom written MATLAB code.

To quantify the mobility of these particles, we calculated the mean squared displacement (MSD) for several trajectories of many cells. Two-dimensional MSD is related to the time window, t, in which it is calculated by MSD(t) ∼ tα, where the exponent, α, is related to the dynamics of the particles. We calculated the MSD across all trajectories of all replicates, with each line on the plot representing an average of 10’s of thousands of single particle trajectories. We found that the MSD is lower for particles in the cytoplasm of freeze-thaw tolerant cells (Figure 3J). This suggests reduced mobility of particles in the cytoplasm of freeze-thaw tolerant cells. From the time and ensemble-averaged MSD, we find α ≈ 0.83 for particles in the cytoplasm of freeze-thaw sensitive cells and α ≈ 0.77 for particles in the cytoplasm of freeze-thaw tolerant cells. The probability density function, p(α), of the exponent, calculated by fitting the time-averaged MSD of a particle (Supplementary Information) to a power law, further reveals a larger fraction of particles exhibiting a significantly restricted motion in the freeze-thaw tolerant cells compared to the WT cells (Figure 3K). We find decreased cytoplasmic mobility and increased confinement of particles in the cytoplasm across all freeze-thaw adapted replicate lines (more details in the Supplementary Information).

Freeze-thaw tolerant cells exhibit quiescence-like properties

The morphological and physical attributes that we described thus-far have been associated with the entry of yeast cells into a quiescent state [38]. It is known that trehalose promotes efficient cellular entry into quiescence and reentry into the cell cycle [37]. Further, higher intracellular trehalose causes a decrease in growth rate of the cell upon re-entry into exponential phase [42, 43]. Higher trehalose levels also lead to a decrease in lag phase duration [16, 44– 46]. Therefore, we measured the population growth characteristics of the control and evolved populations after a freeze-thaw at stationary phase. The optical density (OD) of the cell cultures (absorbance at 600nm) was measured to obtain the relative cell concentrations over the course of growth using a spectrophotometer (Figure 4A).

The growth rates of the populations were calculated for the glucose metabolism phase of exponential growth from the maximal slope of the OD curve. Indeed, we found that the minimum doubling time for the evolved populations was longer than that for the WT populations (Figure 4B). Further, the WT population of cells exhibited a 60% longer lag duration (calculated as time taken to reach the maximal growth rate) than the evolved population of cells (Figure 4C). These results corroborate previous studies that associate higher intracellular trehalose levels with decreased growth rate and lag phase duration and, therefore, transition to a quiescent state.

Selection based on trehalose linked entry into a qui- escence-like state recapitulates the adaptation dynamics.

A Optical density of the cell cultures inoculated into fresh media after a freeze-thaw. B Average doubling time of cells in the population calculated from the growth curves in A. C Average lag phase duration of cells in the population calculated from the growth curves in A. D Survival fraction of cells after freeze-thaw simulated over the course of evolution. E Basal intracellular concentrations of trehalose relative to the control simulated over the course of evolution.

Based on these data, we hypothesized that the freeze-thaw tolerant population is enriched with quiescent cells which leads to a high survival fraction. In this case, selection for quiescence would be equivalent to selection for freeze-thaw tolerance. However, it is not feasible to experimentally select for quiescence; therefore, to investigate this hypothesis, we numerically simulated the adaptation of yeast based on the ability of cells to enter and exit quiescence.

Selection based on trehalose linked entry into a quiescence-like state recapitulates the adaptation dynamics

We developed a population level model with trehalose concentration acting as the sole phenotype under selection, since trehalose levels determine the growth rate, probability of entering quiescence, survival probability after a freezethaw, and lag phase duration before re-entry into the cell cycle [37] (full details of the model are in the Supplementary Information).

Briefly, we start with an initial cell population characterized by a Gaussian distributed phenotypic trehalose levels. The trehalose level for each cell in stationary phase is correlated with its growth rate in log phase, and assume that it directly determines (a) the probability of going into a quiescent state upon encountering starvation conditions [47, 48], (b)the probability of survival after freeze thaw [15, 49–52], and (c) the duration of lag phase upon re-entry into the cell cycle.

In each cycle of growth-freeze-thaw, the population has differing growth rates, and cells with lower trehalose values have a higher growth rate and a larger number of offspring. Each offspring is allowed a “mutation” which acts to shift its stationary state trehalose concentration with respect to its parent. These mutational effects are drawn from a fat-tailed distribution symmetric about zero i.e. most mutations have no effect on the phenotype of the offspring, while the possibility of a few large effect mutations exists. Each cell division uses up a unit of a finite resource, and the population faces starvation when the resource runs out.

Upon starvation, the population differentiates into quiescent and non-quiescent cells. Cells with higher amounts of trehalose have a higher probability of entering a quiescent state. All non-quiescent cells are subsequently assumed to perish when the population is subjected to selection i.e. freeze-thaw. The survival of cells after freeze thaw is determined similarly, using a survival probability that increases with the intracellular trehalose concentration. The population that survives freeze-thaw is diluted by selecting a fraction of the population randomly, which is then allowed to enter growth again. The lag phase of each cell is further determined by its trehalose concentration, with cells with higher trehalose concentrations facing a shorter lag time.

Thus, the states of quiescence and increased survival under freeze thaw cause the population distribution to shift to higher values of mean trehalose. Similarly, cells with higher trehalose have longer periods of growth due to a shorter lag phase, and take over more of the population with each generation of offspring. This effect is diminished slightly by the fact that higher trehalose also leads to a smaller growth rate, but is not strong enough initially to stall the increase in mean trehalose of the population. At larger values of trehalose the effect of a decreasing growth rate becomes stronger, leading to saturation of trehalose levels over the course of evolution. We find that when cells are selected in this way for quiescence, the dynamics of the survival (Figure 4D) and trehalose levels (Figure 4E) of the adapting population qualitatively resemble the experimentally measured adaptation dynamics (Fig 1B,C). We speculate that the more sudden jumps seen in the experimental curves could be due to the larger number of replicates in the simulation than what the experiments than what allow for. Altogether, our results suggest that cellular entry into a quiescence-like state facilitates the adaptation to freeze-thaw.

The mechano-chemical adaptation is convergent

Finally, we ask if the results we describe thus far are a result of the specific selection protocol we used for experimental evolution or if they point to a more general mechanism. Therefore, we sought to investigate the phenotypes emerging from a different freeze-thaw selection regime.

To do this, we developed a selection protocol based on exposing cells to multiple consecutive freeze-thaw cycles before a growth cycle (Figure 5A). First, we note that the survival fraction of the cells decreases exponentially with consecutive freeze-thaw exposures, reducing by about an order of magnitude with the number of consecutive exposures, kFT (Figure 5B). Each point obtained by varying kFT is, therefore, a quantitatively (and qualitatively) different selection regime. Although the results obtained so far have been obtained from the kFT = 1 regime, we now report results from the kFT = 3 selection regime. Interestingly, despite nearly two orders of magnitude reduction in the initial viability, the yeast adapted to the kFT = 3 regime, with survival close to 40%. We find that the population of cells evolved in this regime also exhibits a multifold increase in basal trehalose levels (Figure 5C) and reduced budding fractions (Figure 5D). Furthermore, the MSDs obtained from cells of the evolved population in the kFT = 3 regime fall directly on top of those obtained from the kFT = 1 regime (Figure 5E); the distribution of the MSD exponents, p(α) is also quite similar (Figure 5F). On the basis of these findings, we conclude that the mechano-chemical phenotype promoting cellular entry into a quiescence-like state is a convergent adaptation.

The mechano-chemical adaptation to freeze-thaw tolerance is convergent.

A Experimental design to measure survival after refreezing and for the evolution of refreezing tolerance. B Number of viable cells after consecutive freeze-thaws. C Basal amounts of intracellular trehalose measured in the stationary phase. D Fraction of cells budding measured in the stationary phase. E Mean squared displacement (MSD) of GEMs tracked in the cytoplasm of cells, plotted on logarithmic axes. Inset: MSD plotted on linear axes. F Distribution of the exponent of time, α.

Distinct genetic routes underlie the adaptations

Since we found a convergent phenotype across all replicates of different freeze-thaw regimes, we performed whole genome sequencing to investigate if the adaptations are associated with a particular genotype (or a specific genetic program). We sequenced samples of the evolved and control populations, collected from the middle and end of the evolution experiment, along with the parent strain (full details in the Supplementary Information). The sequenced reads were analyzed post-filtering for the frequency of occurrences of Single Nucleotide Polymorphisms (SNPs), insertions, and deletions in each replicate population. The final list of all such genetic variants obtained for the freeze-thaw tolerant populations, along with their function, are shown in Table V. Each variant is marked for its mutation type and the sample(s) in which it was detected.

We then checked if any variants were present across multiple samples, as this would imply that the mechano-chemical phenotype found across all replicates was a consequence of a specific genotype(s). We found no variants common across all replicates of both kFT = 1 and kFT = 3 selection regimes. It must be noted that mutations in different genes with similar or shared functions, or genes linked in cellular pathways, could lead to the same adaptive phenotype. We investigated this aspect using the curated information available on the Saccharomyces genome database (SGD) on regulatory, metabolic, and physical interactions. We sought to identify a common molecular function or pathway that could have been targeted by different variants across samples. Therefore, we looked for interactions between the different genes and checked if any of the genes are regulators of other genes in the list to understand if there is any complementarity in the genotypes found such that certain replicates have related genotypes. However, no such known cellular network could be found. We also considered changes in gene-copy numbers as a potential basis for observed phenotypes, particularly the increase in intracellular levels of trehalose. Again, we found no gene with a consistent copy-number variation across all replicates (details in the Supplementary Information). These results reveal that the convergent mechano-chemical adaptations seen in this evolution experiment could be achieved with multiple, seemingly disconnected genotypes.

Discussion

Using experimental evolution, we have shown that Saccharomyces cerevisiae can adapt to high freeze rate perturbations in a stereotypical manner, with survival increasing nearly two orders of magnitude over 25 cycles of freezethaw-growth. In all our replicate lines of evolution, the basal intracellular level of trehalose, a sugar speculated to play a physical role in maintaining membrane and cytoplasmic integrity [23, 53, 54], exhibited a significant (three-fold) increase in the adapted cells. The adapted cells also show smaller sizes, higher mass densities, and reduced membrane damage. Using rheological measurements we showed that the adapted cells are stiffer and had lower particle mobility within the cytoplasm. Whole genome sequencing indicated that the increased survival is associated with multiple, distinct genotypes.

Earlier, independent studies have pointed out the role of increased trehalose to confer stress tolerance in yeast, while not connecting the increased trehalose to a broader physiological context (the trehalose was increased artificially via extracellular means) [55]. Studies have also associated cytoplasmic stiffening with transition to dormancy/quiescence in yeast [19, 56] while not looking at such physical changes in an evolutionarily adaptive sense. Further, Sleight et al used experimental evolution with bacteria to shown that microbes can indeed evolve to increase survival to freeze-thaw stress – in this case, the survival increased only 2 fold, compared to the nearly 40 fold in our case – but concluded that “the physiological mechanism for this fitness improvement remains unknown” [29]. To our knowledge, our work is the first study showing the de novo linkage between increased intracellular trehalose, reduced membrane damage, decreased cytoplasmic mobility, increased cellular stiffness and transition to a quiescence-like state.

The high levels of trehalose could play multiple roles in the increased mechanical resilience of the cell. Reduced membrane damage in freeze-thaw tolerant cells could be a result of altered lipid composition, altered cell wall composition and increased trehalose. Several studies show that increased membrane fluidity of cells was associated with higher freeze-thaw tolerance, with trehalose hypothesized to aid in membrane fluidization [29, 34, 57]. In addition, our results on the increased cellular stiffness suggest that the cell wall composition could also be altered, thereby protecting the cell wall associated membrane from undergoing damage.

A major cause of death after freeze-thaw has been suggested to be ice crystal formation [23, 26, 58, 59]. Ice crystals can damage the cell membrane, proteins in the cytoplasm and disrupt the intracellular structure and organisation, thereby hindering the revival of cellular processes upon thawing. A glassy cytoplasm with decreased mobility is a cytoplasmic rheological state that could aid in retaining cellular structure and organisation during freeze-thaw transitions. Indeed such a cytoplasmic state has been implicated in response to various environmental stresses [19, 20, 40], also to be a role of trehalose [54] and has long been associated with freeze-thaw tolerance. We found the de novo encoding of a similar cytoplasmic state with increased intracellular trehalose in the adapted freeze-thaw tolerant cells.

Trehalose has been shown to be a good glass-former, compared to other sugars like glucose and sucrose [60] and has also been found to localise in the cytosol [52]. Several other studies shown that higher intracellular trehalose levels are associated with tolerance to desiccation, heat, ethanol etc [55, 61]. Cakar et al. allowed mutagenic lines of Saccharomyces cerevisiae to adapt to various kinds of stresses and found that freeze-thaw tolerant populations were tolerant to multiple other kinds of stress such as high temperature, ethanol and oxidative stress [62]. The evolution of a mechano-chemically reinforced survival phenotype underscores the role of biophysical constraints in shaping cellular adaptation. Cytoplasmic stiffening and increased trehalose accumulation likely serve as protective mechanisms against mechanical and osmotic damage [19, 20]. This highlights how stress tolerance is not solely dictated by genetic changes but also by modulations in intracellular material properties [13]. Beyond yeast, similar adaptations have been observed in bacterial biofilms and dormant cancer cells [63–65], suggesting broader implications for microbial resilience, antibiotic tolerance, and tumor dormancy.

Interestingly, our sequencing analysis did not reveal any mutations directly related to trehalose synthesis, degradation or transport (Table V). Therefore, this increased concentration of trehalose is not of any direct genetic origin, unlike previous studies that used trehalose synthesis, degradation or transport mutants and related viability after freeze-thaw to intracellular trehalose concentration [30–32].

While freeze-thaw is lethal to most kinds of cells, several cells survive freeze-thaw on ecological and evolutionary timescales. There have been a few previous studies that used experimental evolution to obtain freeze-thaw tolerance in cells. Kwon et al. started with a L. rhamnosus strain that had a 60% survival fraction after freeze-thaw and the adapted strain was found to have a ≈1.5 fold increase in survival fraction over 150 freeze-thaw-growth cycles [57]. Sleight et al. found only a 2.5 times increase in survival fraction of E. coli over 150 freeze-thaw-growth cycles (1000 generations) [29]. Although, these studies are on very different species, they both find cls mutants that increase membrane fluidity to be associated with survival. Since these kinds of stress do not cause a drastic reduction in survival, it could lead to a more gradual phenotypic change as fitness is not affected as much – it must be noted that in our experiments, the survival of wild type Saccharomyces cerevisiae to rapid freeze-thaw is a few percent or lesser. By using freeze-thaw regimes where survival is affected significantly, we found that Saccharomyces cerevisiae adapts to have orders of magnitude higher survival in only ≈25 freeze-thaw-growth cycles (≈100 generations). Our observations of stark phenotypic changes could be because of the strong selection regime in which we operate.

The various cellular attributes that we have discovered in our study correlate with those of quiescent cells. The molecular mechanisms of quiescence entry and exit and the characteristics of the quiescent state are only recently being understood [44, 66] — specifically it has been uncovered that trehalose promotes efficient entry into and exit from quiescence. Our simulations on trehalose-mediated selection for quiescence suggest that an entry to a quiescence-like state, as well as the ease of re-entering the growth cycle (through lag) facilitates the adaptation dynamics we see in the experiments.

Several studies have found that quiescent cells are more stress resilient [67], although the underlying mechanisms underlying have not yet been understood. Specifically, the genetic basis for quiescence is not fully explored. While certain genes involved in the stress response leading to quiescence have been studied through enrichment of quiescent cells in the population [68], these genetic mutations are not present among the adapted populations in our experiments. The emergence of a quiescence-like state in our experimental evolution study, suggests that increased mechanical resilience of a quiescent cell, rather than any specific genetic basis, could be responsible for stress resilience.

By bridging experimental evolution, biophysical characterization, and predictive modeling, our work provides a framework for investigating how cells can rapidly acquire stress tolerance through state transitions rather than relying solely on genetic mutations. Future research should explore the molecular regulation of these adaptive states, including potential epigenetic contributions and metabolic shifts that facilitate quiescence entry and maintenance under extreme conditions [69].

eLife Assessment

This important study demonstrates that yeast populations can rapidly evolve freeze-thaw tolerance by converging on a trehalose-rich, quiescence-like state, illuminating a general physiological route to extreme-stress adaptation. The evidence is solid, combining rigorous experimental-evolution design with multi-scale phenotyping, biophysical measurements, whole-genome sequencing, and quantitative modeling that together support the mechanistic conclusions. Questions about the novelty relative to prior growth/stress tolerance links, the precise genetic versus non-genetic drivers of trehalose up-regulation, and the breadth of independently evolved lines. These are areas for clarification, but these do not substantially weaken the overall contribution.

Reviewer #1 (Public review):

Summary:

This manuscript presents findings on the adaptation mechanisms of Saccharomyces cerevisiae under extreme stress conditions. The authors try to generalize this to adaptation to stress tolerance. A major finding is that S. cerevisiae evolves a quiescence-like state with high trehalose to adapt to freeze-thaw tolerance independent of their genetic background. The manuscript is comprehensive, and each of the conclusions is well supported by careful experiments.

Strengths:

This is excellent interdisciplinary work.

Weaknesses: .

I have questions regarding the overall novelty of the proposal, which I would like the authors to explain.

(1) Earlier papers have shown that loss of ribosomal proteins, that slow growth, leads to better stress tolerance in S. cerevisiae. Given this, isn't it expected that any adaptation that slows down growth would, overall, increase stress tolerance? Even for other systems, it has been shown that slowing down growth (by spore formation in yeast or bacteria/or dauer formation in C. elegans) is an effective strategy to combat stress and hence is a likely route to adaptation. The authors stress this as one of the primary findings. I would like the authors to explain their position, detailing how their findings are unexpected in the context of the literature.

(2) Convergent evolution of traits: I find the results unsurprising. When selecting for a trait, if there is a major mode to adapt to that stress, most of the strains would adapt to that mode, independent of the route. According to me, finding out this major route was the objective of many of the previous reports on adaptive evolution. The surprising part in the previous papers (on adaptive evolution of bacteria or yeast) was the resampling of genes that acquired mutations in multiple replicates of an evolution experiments, providing a handle to understand the major genetic route or the molecular mechanism that guides the adaptation (for example in this case it would be - what guides the over-accumulation of trehalose). I fail to understand why the authors find the results surprising, and I would be happy to understand that from the authors. I may have missed something important.

(3) Adaptive evolution would work on phenotype, as all of selective evolution is supposed to. So, given that one of the phenotypes well-known in literature to allow free-tolerance is trehalose accumulation, I think it is not surprising that this trait is selected. For me, this is not a case of "non-genetic" adaptation as the authors point out: it is likely because perturbation of many genes can individually result in the same outcome - upregulation of trehalose accumulation. Thereby, although the adaptation is genetic, it is not homogeneous across the evolving lines - the end result is. Do the authors check that the trait is actually a non-genetic adaptation, i.e., if they regrow the cells for a few generations without the stress, the cells fall back to being similarly only partially fit to freeze-thaw cycles? Additionally, the inability to identify a network that is conserved in the sequencing does not mean that there is no regulatory pathway. A large number of cryptic pathways may exist to alter cellular metabolic states.

This is a point in continuation of point #2, and I would like to understand what I have missed.

(4) To propose the convergent nature, it would be important to check for independently evolved lines and most probably more than 2 lines. It is not clear from their results section if they have multiple lines that have evolved independently.

(5) For the genomic studies, it is not clear if the authors sequenced a pool or a single colony from the evolved strains. This is an important point, since an average sequence will miss out on many mutations and only focus on the mutations inherited from a common ancestral cell. It is also not clear from the section.

Reviewer #2 (Public review):

Summary:

The authors used experimental evolution, repeatedly subjecting Saccharomyces cerevisiae populations to rapid liquid-nitrogen freeze-thaw cycles while tracking survival, cellular biophysics, metabolite levels, and whole-genome sequence changes. Within 25 cycles, viability rose from ~2 % to ~70 % in all independent lines, demonstrating rapid and highly convergent adaptation despite distinct starting genotypes. Evolved cells accumulated about threefold more intracellular trehalose, adopted a quiescence-like phenotype (smaller, denser, non-budding cells), showed cytoplasmic stiffening and reduced membrane damage, and re-entered growth with shorter lag traits that together protected them from ice-induced injury. Whole-genome sequencing indicated that multiple genetic routes can yield the same mechano-chemical survival strategy. A population model in which trehalose controls quiescence entry, growth rate, lag, and freeze-thaw survival reproduced the empirical dynamics, implicating physiological state transitions rather than specific mutations as the primary adaptive driver. The study therefore concludes that extreme-stress tolerance can evolve quickly through a convergent, trehalose-rich quiescence-like state that reinforces membrane integrity and cytoplasmic structure.

Strengths:

The strengths of the paper are the experimental design, data presentation and interpretation, and that it is well-written.

Weaknesses:

(1) While the phenotyping is thorough, a few more growth curves would be quite revealing to determine the extent of cross-stress protection. For example, comparing growth rates under YPD vs. YPEG (EtOH/glycerol), and measuring growth at 37ºC or in the presence of 0.8 M KCl.

(2) Is GEMS integrated prior to evolution? Are the evolved cells transformable?

(3) From the table, it looks like strains either have mutations in Ras1/2 or Vac8. Given the known requirements of Ras/PKA signaling for the G1/S checkpoint (to make sure there are enough nutrients for S phase), this seems like a pathway worth mentioning and referencing. Regarding Vac8, its emerging roles in NVJ and autophagy suggest another nutrient checkpoint, perhaps through TORC1. The common theme is rewired metabolism, which is probably influencing the carbon shuttling to trehalose synthesis.

Author response:

The following is the authors’ response to the original reviews.

We thank the editor and the reviewers for the detailed and constructive comments. In revising the manuscript we have: (i) clarified what is new relative to prior stress tolerance work, (ii) made explicit that we observe phenotypic convergence without a shared genetic route, (iii) stated upfront that we evolved four independent lines plus two controls, and (iv) corrected figure legends, statistics, and the missing citations. Below we respond point-by-point.

Public Reviews:

Reviewer #1 (Public review):

Summary:

This manuscript presents findings on the adaptation mechanisms of Saccharomyces cerevisiae under extreme stress conditions. The authors try to generalize this to adaptation to stress tolerance. A major finding is that S. cerevisiae evolves a quiescence-like state with high trehalose to adapt to freeze-thaw tolerance independent of their genetic background. The manuscript is comprehensive, and each of the conclusions is well supported by careful experiments.

Strengths:

This is excellent interdisciplinary work.

Weaknesses:

I have questions regarding the overall novelty of the proposal, which I would like the authors to explain.

(1) Earlier papers have shown that loss of ribosomal proteins, that slow growth, leads to better stress tolerance in S. cerevisiae. Given this, isn’t it expected that any adaptation that slows down growth would, overall, increase stress tolerance? Even for other systems, it has been shown that slowing down growth (by spore formation in yeast or bacteria/or dauer formation in C. elegans) is an effective strategy to combat stress and hence is a likely route to adaptation. The authors stress this as one of the primary findings. I would like the authors to explain their position, detailing how their findings are unexpected in the context of the literature.

We agree that the link between slower growth and higher stress tolerance has been well studied. What is distinctive here is that repeated, near-lethal freeze–thaw selected not only for a tolerant/quiescent-like state but also for a shorter lag on re-entry. In this regime of freeze–thaw–regrowth, cells that are tolerant but slow to restart would be outcompeted by naive fast growers. Our quiescence-based selection simulations reproduce exactly this constraint. We have added this explanation to the Results to make clear that the novelty is the co-evolution of a tolerant, trehaloserich state together with rapid regrowth under an alternating regime.

(2) Convergent evolution of traits: I find the results unsurprising. When selecting for a trait, if there is a major mode to adapt to that stress, most of the strains would adapt to that mode, independent of the route. According to me, finding out this major route was the objective of many of the previous reports on adaptive evolution. The surprising part in the previous papers (on adaptive evolution of bacteria or yeast) was the resampling of genes that acquired mutations in multiple replicates of an evolution experiments, providing a handle to understand the major genetic route or the molecular mechanism that guides the adaptation (for example in this case it would be - what guides the overaccumulation of trehalose). I fail to understand why the authors find the results surprising, and I would be happy to understand that from the authors. I may have missed something important.

Our surprise was precisely that we did not see the classical pattern of “phenotypic convergence + repeated mutations in the same locus/module.” All independently evolved lines converged on a trehalose-rich, mechanically reinforced, quiescence-like phenotype, but population sequencing across lines did not reveal a single repeatedly hit gene or small shared pathway, even when we increased selection stringency (1–3 freeze–thaw cycles per round). We have now stated in the manuscript that this decoupling (strong phenotypic convergence, non-overlapping genetic routes) is the central inference: selection is acting on a physiologically defined state that multiple genotypes can reach.

(3) Adaptive evolution would work on phenotype, as all of selective evolution is supposed to. So, given that one of the phenotypes well-known in literature to allow free-tolerance is trehalose accumulation, I think it is not surprising that this trait is selected. For me, this is not a case of ”non-genetic” adaptation as the authors point out: it is likely because perturbation of many genes can individually result in the same outcome - up-regulation of trehalose accumulation. Thereby, although the adaptation is genetic, it is not homogeneous across the evolving lines - the end result is. Do the authors check that the trait is actually a non-genetic adaptation, i.e., if they regrow the cells for a few generations without the stress, the cells fall back to being similarly only partially fit to freeze-thaw cycles? Additionally, the inability to identify a network that is conserved in the sequencing does not mean that there is no regulatory pathway. A large number of cryptic pathways may exist to alter cellular metabolic states.

This is a point in continuation of point #2, and I would like to understand what I have missed.

We agree, and we have removed the wording “non-genetic adaptation.” The evolved populations retain high survival even after regrowth for ≥25 generations without freeze–thaw, so the adaptation is clearly genetically maintained. What our data show is that there is no single genetic route to the shared phenotype; different mutations can all drive cells into the same trehalose-rich, quiescencelike, mechanochemically reinforced state. We now describe this as “genetic diversification with phenotypic convergence.”

(4) To propose the convergent nature, it would be important to check for independently evolved lines and most probably more than 2 lines. It is not clear from their results section if they have multiple lines that have evolved independently.

We indeed evolved four independent lines and maintained two independent controls. We have added this information at the start of the Results so that the level of replication is immediately clear.

(5) For the genomic studies, it is not clear if the authors sequenced a pool or a single colony from the evolved strains. This is an important point, since an average sequence will miss out on many mutations and only focus on the mutations inherited from a common ancestral cell. It is also not clear from the section.

We sequenced population samples from the evolved lines. Our specific question was whether independently evolved lines would show the same high-frequency genetic solution, as is often seen in parallel evolution. Pool sequencing may under-sample rare/private variants, but it is appropriate for detecting such shared, high-frequency routes — and we do not find any. We have clarified this rationale in the Methods/Results.

Reviewer #2 (Public review):

Summary:

The authors used experimental evolution, repeatedly subjecting Saccharomyces cerevisiae populations to rapid liquid-nitrogen freeze-thaw cycles while tracking survival, cellular biophysics, metabolite levels, and whole-genome sequence changes. Within 25 cycles, viability rose from ~2 % to ~70 % in all independent lines, demonstrating rapid and highly convergent adaptation despite distinct starting genotypes. Evolved cells accumulated about threefold more intracellular trehalose, adopted a quiescence-like phenotype (smaller, denser, non-budding cells), showed cytoplasmic stiffening and reduced membrane damage, and re-entered growth with shorter lag traits that together protected them from ice-induced injury. Whole-genome sequencing indicated that multiple genetic routes can yield the same mechano-chemical survival strategy. A population model in which trehalose controls quiescence entry, growth rate, lag, and freeze-thaw survival reproduced the empirical dynamics, implicating physiological state transitions rather than specific mutations as the primary adaptive driver. The study therefore concludes that extreme-stress tolerance can evolve quickly through a convergent, trehalose-rich quiescence-like state that reinforces membrane integrity and cytoplasmic structure.

Strengths:

The strengths of the paper are the experimental design, data presentation and interpretation, and that it is well-written.

(1) While the phenotyping is thorough, a few more growth curves would be quite revealing to determine the extent of cross-stress protection. For example, comparing growth rates under YPD vs. YPEG (EtOH/glycerol), and measuring growth at 37ºC or in the presence of 0.8 M KCl.

We thank the referee for the interesting suggestions. However, growth rates alone may be difficult to interpret since WT strains also show different growth rates under these conditions. Therefore, comparing the relative fitness or survival of the evolved strains versus the WT under these stresses would be more informative. In the present study we limited growth/survival measurements to what was needed to parameterize the adaptation model in YPD under the freeze–thaw regime. We have now added a statement in the Discussion that, given the shared trehalose/mechanical mechanism, such cross-stress assays are an expected and straightforward follow-up.

(2) Is GEMS integrated prior to evolution? Are the evolved cells transformable?

Yes. GEMs were integrated prior to evolution, because the non-integrated evolved population showed low transformation efficiency, likely due to altered cell-wall properties.

(3) From the table, it looks like strains either have mutations in Ras1/2 or Vac8. Given the known requirements of Ras/PKA signaling for the G1/S checkpoint (to make sure there are enough nutrients for S phase), this seems like a pathway worth mentioning and referencing. Regarding Vac8, its emerging roles in NVJ and autophagy suggest another nutrient checkpoint, perhaps through TORC1. The common theme is rewired metabolism, which is probably influencing the carbon shuttling to trehalose synthesis.

We appreciate the reviewer’s suggestion to consider pathways like Ras/PKA (linked to Ras1/2) and autophagy/TORC1 (linked to Vac8) as potential upstream modulators. While these pathways are involved in nutrient sensing and metabolic regulation, we choose not to emphasize them specifically. This is because (i) some evolved lines lack Ras1/2 or Vac8 variants, and (ii) none of the variants lies directly in trehalose synthesis/degradation pathways. Furthermore, direct links to trehalose accumulation are not well established for these specific variants in this context, and pathways like Ras are global regulators with broad effects. Together with the strongly convergent phenotype, this supports our main inference that multiple genetic/metabolic routes can feed into the same trehalose-rich, mechanochemically reinforced, quiescence-like state. We have added a note in the discussion regarding metabolic rewiring and trehalose.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

Generally, the results sections should have more details. The figures should be corrected, and the legends should be checked for correctness. The manuscript seems to have been assembled in haste?

We have expanded the relevant Results subsections with one-sentence motivations (why each measurement was performed) and we have corrected the figure legends for ordering and consistency.

Figure 3: It will be good to have the correct p-values on the figure itself. P-values are typically less than 1, unless there is some special method (here the values presented are , etc). Please explain how the P-values were obtained in the figure legend itself.

Figure 3 now shows the actual p-values. The legend specifies the details and the sample sizes used.

Figure 5: It is not clear what the error bars show in 5B, E (different evolved population/ clones/ cells?). All the figure legends are mixed up, please correct them. It is difficult to follow the paper.

Figure 5 legends now state clearly what the error bars represent (biological replicates) and which panels are from single-cell measurements. We have checked the panel lettering and legend order for consistency with the flow of the main text.

Reviewer #3 (Recommendations for the authors):

Overall, the paper is outstanding, well-written, and insightful.

A point to address is that there are missing citations on lines 60, 91.

We have added the missing citations at both locations. We apologize for the omission, which was due to a compilation error. This error has been fixed, and the bibliography has been corrected (now containing 74 references).

Version 2

Abstract

Extreme stress, such as freeze-thaw, poses a severe challenge to many organisms, but the mechanisms underlying their adaptation to survive such stress remain elusive. Here, we show that Saccharomyces cerevisiae can rapidly evolve freeze-thaw tolerance through a physiological state transition, with survival increasing nearly two orders of magnitude from ≈2% to ≈70% in about 25 cycles of stress exposure. Evolved yeast cells exhibit a quiescence-like state, characterized by altered cellular physiology: increased intracellular trehalose accumulation, reduced membrane damage, cytoplasmic stiffening and an exit from a proliferative cycle. This mechano-chemically reinforced survival strategy emerges across independent evolutionary lines despite distinct genetic backgrounds, suggesting a convergent mechanism of adaptation. By integrating experimental evolution, biophysical measurements, genomic analysis, and a quantitative model that captures the adaptation dynamics, we reveal that stress tolerance can arise via a potentially generalizable, physiologically mediated adaptation strategy. These findings provide new insights into microbial survival under extreme conditions and suggest broader implications for cellular stress responses beyond yeast.

Introduction

Cells across many domains of life survive extreme environmental perturbations, including complete desiccation and freezing, which can cause severe cellular damage. Cellular damage occurs via many mechanisms such as protein denaturation [1, 2], membrane rupture [3, 4], and oxidative stress [5]. To survive such conditions, organisms have evolved diverse adaptive mechanisms [6, 7]. Classical evolutionary models primarily emphasize genetic mutations as the basis for the evolution of such adaptations [8–10]. However, recent evidence suggests that cells can also adapt via phenotypic plasticity and physiological state transitions, allowing rapid, non-genetic survival advantages [11–16].

One widely recognized mechanism of stress resistance is the accumulation of protective molecules such as trehalose, which stabilizes proteins and membranes during dehydration and freezing [14, 17, 18]. Additionally, molecules such as trehalose have also been associated with transition to quiescence, a reversible state of metabolic suppression, to endure unfavorable conditions, in many microbes and multicellular organisms. Among many other characteristics, quiescence is characterized by increased cytoplasmic viscosity, enhanced resistance to oxidative damage, and improved long-term viability [19–22]. In microbes, quiescence has been linked to persistence under starvation and antimicrobial stress, while in multicellular organisms, it plays a role in stem cell maintenance and tumor dormancy [23–25].

Here, we investigate whether repeated exposure to extreme freeze-thaw stress selects for a quiescence-like adapted state in Saccharomyces cerevisiae. Using experimental evolution, we subjected yeast populations to repeated cycles of rapid freezing and thawing, a severe mechano-chemical perturbation that initially results in very low survival (<2%) [26, 27]. Strikingly, within ∼25 selection cycles, survival increased nearly 40-fold. The adapted yeast cells exhibit an increased prevalence of a quiescence-like state, characterized by altered cellular physiology: increased intracellular trehalose accumulation, reduced membrane damage, cytoplasmic stiffening and an exit from a proliferative cycle. Whole-genome sequencing revealed that the adaptation was not driven by unique genetic pathways but rather by multiple, distinct genetic backgrounds, all converging on the same mechano-chemically reinforced survival phenotype. Our findings suggest that a quiescence-like adaptation represents a generalizable stress tolerance strategy. These results raise intriguing questions about whether similar physiological state transitions contribute to microbial persistence in other kinds of fluctuating environments [28] and whether such mechanisms can be harnessed for industrial and medical applications, including bio-preservation and cryo-tolerance. By demonstrating that cells can evolve resilience via a physiologically mediated mechanism that operates alongside genetic factors, our work expands the understanding of adaptive evolution and stress biology beyond classical mutation-driven models [29].

Results

Saccharomyces cerevisiae adapts to freeze-thaw stress

We first establish an experimental protocol in order to perform the selection (Methods; Figure 1A). It is known that the survival of Saccharomyces cerevisiae upon freeze-thaw depends on the rate of freezing and thawing. Survival has been found to be minimal when it undergoes a fast freeze (> 10 K/min) to −196◦C [30]. We chose this regime to work in to possibly achieve the most significant increase in survival after adaptation. Before proceeding, we determined the effect of duration at −196◦C on survival by exposing the Wild Type - CEN.PK a/ α (WT) stationary phase cell cultures to a fast freeze (> 10K/min) to −196◦C and allowed them to remain at −196◦C for varying lengths of time before thawing. Freezing was done by immersing 500 µl of stationary phase culture in 1.5 mL tubes in liquid nitrogen (−196◦C) and the cultures were then thawed by immersing them in 25◦C water for 15 minutes. We note that the survival fraction of yeast does not vary with duration at −196◦C (Supplementary Information). Therefore, for experimental convenience, we used a freeze duration of 15 minutes and selected for freeze-thaw tolerance at high cooling rates (> 10 K/min).

Experimental evolution can be used to increase survival of yeast under freeze-thaw stress.

A Experimental design of selection for freeze-thaw tolerance. B Survival fraction of cells after freeze-thaw over the course of evolution. Each point on the thick purple line is the average of four biological replicates and two technical replicates for each biological replicate. Error bars indicate standard deviation. Each light coloured line represents a biological replicate. The evolution was performed thrice and the data from one evolution experiment is shown here. Inset: Average number of viable cells after every freeze- thaw-growth cycle. Purple - evolving population. Pink - control population. C Basal intracellular concentrations of trehalose relative to the control measured over the course of evolution. Each point on the thick purple line is the average of 4 biological replicates and 12 technical replicates for each biological replicate. Error bars indicate standard deviation. Each light coloured line represents a biological replicate. This measurement was made during one evolution experiment.

To start the experiments, a single yeast colony was picked from an agar plate and grown as liquid culture to stationary phase. This starter culture was divided into six parts — two for controls and four biological replicates for selection. The four biological replicates underwent freeze-thaw and the survival fraction was then measured by colony forming units (CFUs). The four populations that underwent freeze-thaw were independently grown in fresh medium until stationary phase (24 hours, ≈4 generations) and were independently put through freeze-thaw again. This cycle of freeze-thaw followed by growth was repeated several times until the survival fraction stopped increasing. For controls, the same protocol was followed without the freeze-thaw procedure.

Under this selection protocol, the survival fraction increased from ≈2% up to an average of ≈70%, with one of the replicates as high as 85%, in only a span of ≈25 freeze-thaw cycles. This shows that the yeast cells adapt quite rapidly to freeze-thaw in this regime of very low initial survival rates (i.e. ≈2%) and the dynamics are remarkably similar across independent evolution lines (Figure 1B). Next, we characterized the differences in the evolved strain with respect to the WT.

Freeze-thaw tolerant populations of cells have higher basal intracellular levels of trehalose

Trehalose is a sugar containing two α-glucose units with a 1,1-glycosidic bond between them. Several studies have found a strong correlation between intracellular trehalose concentration and survival after freeze-thaw [31, 32]. Chen et al. showed that trehalose was sufficient to confer freeze-thaw tolerance when cooled slowly to −80◦C [33]. These studies either measure the amount of trehalose in existing freeze-thaw tolerant strains or the survival in trehalose accumulated mutant strains. Motivated by these studies, we measured the amounts of intracellular trehalose at the stationary phase to calculate the basal levels of trehalose present in cells, over the course of adaptation to freeze-thaw. Cell colonies were collected and weighed from plates used to measure survival using CFUs 48 hours after thawing. The concentration of trehalose was then measured using a previously described protocol [34] (Methods). We found that the amount of basal intracellular trehalose systematically increased over the course of adaptation to freeze-thaw, exhibiting similar dynamics to those of the cellular survival levels (Figure 1C). Basal levels of trehalose increased nearly 3 fold over the course of the selection.

Previous studies suggest that trehalose might play a role in reducing membrane damage [35], which has been implicated as the primary cause of death after freeze-thaw [36–38]. Therefore, we next turned our attention to investigating if the freeze-thaw tolerant cells have a membrane-damage related adaptation.

Freeze-thaw tolerant cells undergo reduced membrane damage

The presence or absence of membrane damage was measured by staining the cells with two dyes — 5-carboxyfluorescein diacetate (5-CFDA) and Propidium Iodide (PI) — and measuring the intensities of fluorescence of both dyes using flow cytometry (Methods). The fluorescent form of CFDA leaks out of a cell quickly if it is membrane damaged. PI permeates cells with damaged membranes. Taken together, these dyes allow a clear distinction between membrane-damaged cells and cells with intact membranes, showing up as two distinct clusters in the data (Figure 2A,B). Each point on these plots represents a single cell. We compute a single metric that combines this information to estimate the total membrane damage (Methods), which clearly indicates that a smaller fraction of freeze-thaw tolerant cells undergoes membrane damage after freeze-thaw (Figure 2C). Further, we sought to understand the effect of membrane damage on cellular survival. We used fluorescence-activated cell sorting (FACS) to sort the cells into “membrane-damaged” and “membrane-intact” fractions. We then measured the number of colony forming units (CFUs) within each sorted fraction to quantify their viability (detailed description in Supplementary Information). We found that the survival fraction of freeze-thaw tolerant cells within the membrane-damaged sub-population exceeded the wild type by about two orders of magnitude (Figure 2D). Similarly, the viability of freeze-thaw tolerant cells with intact membranes was an order of magnitude higher than the corresponding wild type population.

Freeze-thaw tolerant cells undergo reduced membrane damage and survive membrane damage.

A Flow cytometry measurements on control cells. Each point represents a single cell and the plot contains data for 10000 cells. The x-axis is the fluorescence intensity in the 5-CFDA channel. The y-axis is the fluorescence intensity in the propidium iodide channel. The contour lines represent changes in density of cells, starting from 17 with increments of 17. The cluster on top comprises membrane damaged cells and the cluster at the bottom comprises cells with intact membranes. B Flow cytometry measurements on freeze-thaw tolerant cells. Each point represents a single cell and the plot contains data for 10000 cells. The x-axis is the fluorescence intensity in the 5-CFDA channel. The y-axis is the fluorescence intensity in the propidium iodide channel. The contour lines represent changes in density of cells, starting from 17 with increments of 17. The cluster on top comprises membrane damaged cells and the cluster at the bottom comprises cells with intact membranes. C The 2-dimensional data from A and B are used to compute a 1-dimensional distribution of “membrane damage” in cells. Area under the normalized curves: Evolved (Membrane intact) = 0.6347, Evolved (Membrane damaged) = 0.3127, Control (Membrane intact) = 0.2331, Control (Membrane damaged) = 0.7597. D Viability of cells measured after sorting the two clusters identified from the data in A and B. A 0.3% sorting error was measured by sorting beads. The entire experiment described in this figure was performed four times and the data from one experiment is shown here.

These data indicate that: (i) a significant fraction of cellular death following freeze-thaw is unrelated to membrane damage and (ii) the evolved cells have mechanisms that can aid in survival following membrane damage. This suggests cellular mechanisms and attributes beyond just the membrane protective roles of trehalose.

Increased intracellular trehalose is a defining characteristic of cellular quiescence — a temporary non-proliferating state — during which cells are tolerant to a range of stress conditions [39]. Therefore, we next investigated if the freeze-thaw tolerant cells exhibit traits similar to those of quiescent cells [40]. We started by probing the morphological and physical traits of freeze-thaw tolerant cells in the stationary phase – the point in their population growth stage at which the freeze-thaw stress was imposed.

Freeze-thaw tolerant cells are smaller, denser, and exhibit reduced budding

Quiescent cells have been shown to be smaller and denser /[41]. Therefore, to investigate if there could be any morphological changes associated with the increased survival of cells, we imaged cells of the wild-type yeast and the freeze-thaw tolerant yeast at stationary phase to measure their size (Figure 3A). We observed a two-fold reduction in cell area (Figure 3B), which corresponds to an estimated six-fold volume reduction in freeze-tolerant cells. A decrease in cell volume could be due to a decrease in its mass and/or increase in cell density. By measuring the single cell mass density of cells at stationary phase using a density gradient column, we found that the freeze-thaw tolerant cells had a significantly higher cell density (Figure 3C).

Cells of the population adapted to survive freeze and thaw are smaller, denser and exhibit reduced budding.

A Brightfield image of cells. Scale bar - 10µm. B Projected area of cells measured from images as shown in A. On each box, the central mark indicates the median, and the bottom and top edges of the box indicate the 25th and 75th percentiles, respectively. The whiskers extend to the most extreme data points not considered outliers, and the outliers are plotted individually as points scattered in the x-direction. Sample size (number of cells measured): Control = 1107, Evolved = 4281, Welch’s t-test: P = 1.3 × 10−213. This experiment was performed twice and the data from one experiment for one evolved population is shown here. C Single cell mass density measured using a percoll gradient column. On each box, the central mark indicates the median, and the bottom and top edges of the box indicate the 25th and 75th percentiles, respectively. The whiskers extend to the most extreme data points not considered outliers (there are no outliers). Welch’s t-test: P = 3.2 × 10−7. This experiment was performed twice and the data from both experiments for eight technical replicates of one biological replicate was pooled. D Fraction of cells budding in stationary phase measured from images as shown in A. On each box, the central mark indicates the median, and the bottom and top edges of the box indicate the 25th and 75th percentiles, respectively. The whiskers extend to the most extreme data points not considered outliers (there are no outliers). Sample size (number of cells measured): Control = 309, Evolved = 422, Welch’s t-test: P = 2.59 × 10−6. This experiment was performed twice and the data from one experiment for one evolved population is shown here. E Distribution of normalized fluorescence intensities of DNA in cells stained with SYTOX Green measured using flow cytometry. Area under the normalized curves: Evolved (Single DNA intensity) = 0.8726, Evolved (Double DNA intensity) = 0.1274, Control (Single DNA intensity) = 0.8330, Control (Double DNA intensity) = 0.1670. Sample size (number of cells measured): Control = 199682, Evolved = 399113. This experiment was performed twice and the pooled data for four biological replicates from one experiment is shown here. F Design of the microfluidic channel through which cells are flowed to measure deformation with an image of cells leaving the microfluidic channel. Scale bar - 15µm. G Eccentricity and area of cells measured in the narrow channel. Flow rate: Control - 5.5cm/s, Evolved - 5.6cm/s. Sample size (number of cells measured): Control = 5825, Evolved = 6192. This experiment was performed four times for one evolved population and the data from one experiment is shown here. H Fraction of intact cells over time measured by turbidity of cells placed in water after Zymolyase treatment. This experiment was performed five times for one evolved population and three technical replicates and the pooled data for all five experiments is shown here. I Image of Genetically Encoded Multimers (GEMs) in the cytoplasm of cells. Scale bar - 5µm. J Mean squared displacement (MSD) of GEMs tracked in the cytoplasm of cells, plotted on logarithmic axes. Inset: MSD plotted on linear axes. More than 10,000 trajectories were used to calculate the average. K Distribution of the exponent of time, α. More than 10,000 trajectories were used to plot the histogram.

From the cellular morphologies, we also found a reduction in the incidence of budding in the evolved cell population (Figure 3D), suggesting that they stopped doubling before reaching stationary phase and could have exited the proliferating phase of the cell cycle - a sign of quiescence entry. This is also corroborated by a measurement of total amount of cellular DNA — the evolved cell population has reduced fraction of dual DNA copies (Figure 3E), suggestive of quiescence.

The increased cellular mass density suggests that there could be an associated increase in molecular crowding in the cytoplasm, thereby reducing cytoplasmic mobility and increasing cellular stiffness. Since we also observed reduced membrane damage, we investigated if the freeze-thaw tolerant cells are more mechanically resilient.

Freeze-thaw tolerant cells are mechanically stiffer

Yeast cells are naturally ellipsoid due to the presence of a stiff cell wall. Exposure to Zymolyase, a cell-wall digesting enzyme, leads, over time, to the formation “spheroplasts”, spherical cells bound by a lipid membrane. We used these Zymolyase treated cells to quantify their stiffness, by employing deformability cytometry a method previously described by Otto et al. [42]. Briefly, the cytometry is performed by suspending the Zymolyase treated cells in a viscous buffer and flowing them through a narrow microfluidic channel. The resultant shear causes cellular deformation, which is measured using high-speed imaging (Figure 3F). The deformation of cells due to shear is quantified by their eccentricity in relation to their size. For a given cell size, lower cell stiffness results in a higher deformation and, therefore, higher eccentricity. From such a measurement, we found two clusters of freeze-thaw tolerant cells with a majority of cells minimally deformed, compared to the freeze-thaw sensitive (WT) cells (Figure 3G). The presence of two clusters indicates increased cellular stiffness even after the long Zymolyase treatment. Munder et al. found that cells that had increased cytoplasmic stiffness did not relax into spheres after treatment with Zymolyase [19]. Hence, the relaxation of freeze-thaw tolerant cells from ellipsoidal to spherical upon treatment with Zymolyase could be affected either by their cytoplasmic stiffness or their cell wall resilience due to a varied cell wall composition that is not affected by Zymolyase to the same extent. Therefore, we further investigated the role of these two factors – cell wall resilience and cytoplasmic stiffness – in increased cellular stiffness.

In yeast, the cell wall maintains the structure of the cell. When the cell wall is removed and cells are placed in a hypo-osmotic environment, cells lyse as they cannot regulate the amount of water permeating the cell. Therefore, to investigate the presence of the cell wall after digestion with Zymolyase, we exposed cells to Zymolyase for varying lengths of time in an appropriate buffer followed by a hypo-osmotic environment i.e. pure water (Supplementary Information). Cells lacking cell-wall lyse in water. So, we measured the turbidity of the solution after placing the cells in water as a measure of the fraction of intact cells. A significant fraction of the freeze-thaw tolerant cells do not lyse even after exposure to Zymolyase for long periods of time, compared to the duration taken for most of the freeze-thaw sensitive cells to lyse (Figure 3H). This suggests three possibilities - (a) the freeze-thaw tolerant cells have an altered cell wall composition that is not easily digested by Zymolyase, (b) the cells have an altered membrane structure (as seen above) that does not allow water to enter the cell even after cell wall digestion, or (c) an increased cytoplasmic stiffness that does not allow cells to relax after cell wall digestion and hence the cellular components do not spread into solution enough to change the turbidity of the solution.

These results suggest that cells of the adapted population have increased mechanical (structural) resilience associated with modified cell wall or cytoplasmic properties. Next, we specifically investigated the cytoplasmic rheology to understand the nature of the increased cellular stiffness.

Freeze-thaw tolerant cells exhibit reduced cytoplasmic mobility

To study the cytoplasmic mobility, we tracked the intracellular motion of engineered fluorescent particles (41 nm diameter) (Figure 3I). Briefly, to produce these fluorescent particles, we transformed the wild type CEN.PK strain to express genetically encoded multimers (GEMs) using a plasmid developed by Delarue et al. [43]. Delarue et al. replaced the INO4 promoter with the monomeric fluorophore sequence fused with the INO4 promoter and selected for the integrants using uracil auxotrophy. We inserted the monomeric fluorophore fused with the INO4 promoter at the HO locus in a prototroph which does not affect growth characteristics significantly in a diploid strain [44]. The plasmid also contained a KanMX marker for selecting the integrants using resistance to G418 (full details in the Supplementary Information). Evolution was performed on the wild type strain transformed to express GEMs. We imaged the GEMs expressed in the cells using total internal reflection fluorescence (TIRF) microscopy, used image analysis to track the particles and analysed the trajectory dynamics using custom written MATLAB code.

To quantify the mobility of these particles, we calculated the mean squared displacement (MSD) for several trajectories of many cells. Two-dimensional MSD is related to the time window, t, in which it is calculated by MSD(t) ∼ tα, where the exponent, α, is related to the dynamics of the particles. We calculated the MSD across all trajectories of all replicates, with each line on the plot representing an average of 10’s of thousands of single particle trajectories. We found that the MSD is lower for particles in the cytoplasm of freeze-thaw tolerant cells (Figure 3J). This suggests reduced mobility of particles in the cytoplasm of freeze-thaw tolerant cells. From the time and ensemble-averaged MSD, we find α ≈ 0.83 for particles in the cytoplasm of freeze-thaw sensitive cells and α ≈ 0.77 for particles in the cytoplasm of freeze-thaw tolerant cells. The probability density function, p(α), of the exponent, calculated by fitting the time-averaged MSD of a particle (Supplementary Information) to a power law, further reveals a larger fraction of particles exhibiting a significantly restricted motion in the freeze-thaw tolerant cells compared to the control cells (Figure 3K). We find decreased cytoplasmic mobility and increased confinement of particles in the cytoplasm across all freeze-thaw adapted replicate lines (more details in the Supplementary Information).

Freeze-thaw tolerant cells exhibit quiescence-like properties

The morphological and physical attributes that we described thus-far have been associated with the entry of yeast cells into a quiescent state [40]. It is known that trehalose promotes efficient cellular entry into quiescence and re-entry into the cell cycle [39]. Further, higher intracellular trehalose causes a decrease in growth rate of the cell upon re-entry into exponential phase [45, 46]. Higher trehalose levels also lead to a decrease in lag phase duration [16, 47–49]. Therefore, we measured the population growth characteristics of the control and evolved populations after a freeze-thaw at stationary phase. The optical density (OD) of the cell cultures (absorbance at 600nm) was measured to obtain the relative cell concentrations over the course of growth using a spectrophotometer (Figure 4A).

Selection based on trehalose linked entry into a quiescence-like state recapitulates the adaptation dynamics.

A Optical density of the cell cultures inoculated into fresh media after a freeze-thaw. B Average doubling time of cells in the population calculated from the growth curves in A. C Average lag phase duration of cells in the population calculated from the growth curves in A. D Survival fraction of cells after freeze-thaw simulated over the course of evolution. Each light coloured line represents an iteration of the simulation. The thick purple line indicates the average for 25 iterations. Error bars indicate standard deviation. E Basal intracellular concentrations of trehalose relative to the control simulated over the course of evolution. Each light coloured line represents the average of the population in an iteration of the simulation. The thick purple line indicates the average for 25 iterations. Error bars indicate standard deviation of 25 iterations.

The growth rates of the populations were calculated for the glucose metabolism phase of exponential growth from the maximal slope of the OD curve. Indeed, we found that the minimum doubling time for the evolved populations was longer than that for the control populations (Figure 4B). Further, the control population of cells exhibited a 60% longer lag duration (calculated as time taken to reach the maximal growth rate) than the evolved population of cells (Figure 4C). Therefore, we observe co-evolution of a freeze-thaw tolerant, trehalose-rich state together with rapid regrowth under a stress-growth alternating regime. These results corroborate previous studies that associate higher intracellular trehalose levels with decreased growth rate and lag phase duration and, therefore, transition to a quiescent state.

Based on these data, we hypothesized that the freeze-thaw tolerant population is enriched with quiescent cells which leads to a high survival fraction. In this case, selection for quiescence would be equivalent to selection for freeze-thaw tolerance. However, it is not feasible to experimentally select for quiescence; therefore, to investigate this hypothesis, we numerically simulated the adaptation of yeast based on the ability of cells to enter and exit quiescence.

Selection based on trehalose linked entry into a quiescence-like state recapitulates the adaptation dynamics

We developed a population level model with trehalose concentration acting as the sole phenotype under selection, since trehalose levels determine the growth rate, probability of entering quiescence, survival probability after a freeze-thaw, and lag phase duration before re-entry into the cell cycle [39] (full details of the model are in the Supplementary Information).

Briefly, we start with an initial cell population characterized by a Gaussian distributed phenotypic trehalose levels. The trehalose level for each cell in stationary phase is correlated with its growth rate in log phase, and assume that it directly determines (a) the probability of going into a quiescent state upon encountering starvation conditions [50, 51], (b) the probability of survival after freeze thaw [15, 52–55], and (c) the duration of lag phase upon re-entry into the cell cycle.

In each cycle of growth-freeze-thaw, the population has differing growth rates, and cells with lower trehalose values have a higher growth rate and a larger number of offspring. Each offspring is allowed a “mutation” which acts to shift its stationary state trehalose concentration with respect to its parent. These mutational effects are drawn from a fat-tailed distribution symmetric about zero i.e. most mutations have no effect on the phenotype of the offspring, while the possibility of a few large effect mutations exists. Each cell division uses up a unit of a finite resource, and the population faces starvation when the resource runs out.

Upon starvation, the population differentiates into quiescent and non-quiescent cells. Cells with higher amounts of trehalose have a higher probability of entering a quiescent state. All non-quiescent cells are subsequently assumed to perish when the population is subjected to selection i.e. freeze-thaw. The survival of cells after freeze thaw is determined similarly, using a survival probability that increases with the intracellular trehalose concentration. The population that survives freeze-thaw is diluted by selecting a fraction of the population randomly, which is then allowed to enter growth again. The lag phase of each cell is further determined by its trehalose concentration, with cells with higher trehalose concentrations facing a shorter lag time.

Thus, the states of quiescence and increased survival under freeze thaw cause the population distribution to shift to higher values of mean trehalose. Similarly, cells with higher trehalose have longer periods of growth due to a shorter lag phase, and take over more of the population with each generation of offspring. This effect is diminished slightly by the fact that higher trehalose also leads to a smaller growth rate, but is not strong enough initially to stall the increase in mean trehalose of the population. At larger values of trehalose the effect of a decreasing growth rate becomes stronger, leading to saturation of trehalose levels over the course of evolution. We find that when cells are selected in this way for quiescence, the dynamics of the survival (Figure 4D) and trehalose levels (Figure 4E) of the adapting population qualitatively resemble the experimentally measured adaptation dynamics (Fig 1B,C). We speculate that the more sudden jumps seen in the experimental curves could be due to the larger number of replicates in the simulation than what the experiments than what allow for. Altogether, our results suggest that cellular entry into a quiescence-like state facilitates the adaptation to freeze-thaw.

The mechano-chemical adaptation is convergent

Finally, we ask if the results we describe thus far are a result of the specific selection protocol we used for experimental evolution or if they point to a more general mechanism. Therefore, we sought to investigate the phenotypes emerging from a different freeze-thaw selection regime.

To do this, we developed a selection protocol based on exposing cells to multiple consecutive freeze-thaw cycles before a growth cycle (Figure 5A). First, we note that the survival fraction of the cells decreases exponentially with consecutive freeze-thaw exposures, reducing by about an order of magnitude with the number of consecutive exposures, kFT (Figure 5B). Each point obtained by varying kFT is, therefore, a quantitatively (and qualitatively) different selection regime. Although the results obtained so far have been obtained from the kFT = 1 regime, we now report results from the kFT = 3 selection regime. Interestingly, despite nearly two orders of magnitude reduction in the initial viability, the yeast adapted to the kFT = 3 regime, with survival close to 40%. We find that the population of cells evolved in this regime also exhibits a multifold increase in basal trehalose levels (Figure 5C) and reduced budding fractions (Figure 5D). Furthermore, the MSDs obtained from cells of the evolved population in the kFT = 3 regime fall directly on top of those obtained from the kFT = 1 regime (Figure 5E); the distribution of the MSD exponents, p(α) is also quite similar (Figure 5F). On the basis of these findings, we conclude that the mechano-chemical phenotype promoting cellular entry into a quiescence-like state is a convergent adaptation.

The mechano-chemical adaptation to freeze-thaw tolerance is convergent.

A Experimental design to measure survival after refreezing and for the evolution of refreezing tolerance. B Number of viable cells after consecutive freeze-thaws. Error bars indicate standard deviation. This experiment was performed five times with two technical replicates and the pooled data from all experiments is shown here. The point marked in blue denotes the selection pressure for (kF T = 1) evolution. The point marked in green denotes the selection pressure for (kF T = 3) evolution. Basal amounts of intracellular trehalose measured in the stationary phase. On each box, the central mark indicates the median, and the bottom and top edges of the box indicate the 25th and 75th percentiles, respectively. The whiskers extend to the most extreme data points not considered outliers (there are no outliers). P (kF T = 1) = 7.1 ×10−8, P (kF T = 3) = 1.6 ×10−7. DFraction of cells budding measured in the stationary phase. On each box, the central mark indicates the median, and the bottom and top edges of the box indicate the 25th and 75th percentiles, respectively. The whiskers extend to the most extreme data points not considered outliers, and the outliers are plotted individually as points scattered in the x-direction. This experiment was performed once and the data for one evolved population and six technical replicates is shown here. Sample size: Control (kF T = 1) = 309, Evolved (kF T = 1) = 422, P (kF T = 1) = 2.5 × 10−6, Control (kF T = 3) = 357, Evolved (kF T = 3) = 459, P (kF T = 3) = 8.1 × 10−5. E Mean squared displacement (MSD) of GEMs tracked in the cytoplasm of cells, plotted on logarithmic axes. Inset: MSD plotted on linear axes. More than 10,000 trajectories were used to calculate the average. F Distribution of the exponent of time, α. More than 10,000 trajectories were used to plot the histogram.

Distinct genetic routes underlie the adaptations

Since we found a convergent phenotype across all replicates of different freeze-thaw regimes, we sought to understand if the phenotype is a physiological response with epigenetic inheritance or whether it was more likely to be genetically inherited. We propagated the evolved lines without freeze-thaw stress for more than 25 generations and found that survival fraction did not reduce significantly (details in the Supplementary Information), suggesting genetic inheritance. We then performed whole genome sequencing to investigate if the adaptations are associated with a particular genotype (or a specific genetic program).

DNA from cells of samples of the evolved and control populations, collected from the middle and end of the evolution experiment, along with the parent strain was isolated. Sequencing was performed on whole individual populations for the evolved and control replicates and from a single clone for the parent strain. Since we were interested in shared and high-frequency genetic routes, we performed population sequencing even though it might result in under-sampling of rare variants. The sequenced reads were trimmed and filtered based on quality and then mapped to a reference genome. The mapped reads were analyzed post-filtering for the frequency of occurrences of Single Nucleotide Polymorphisms (SNPs), insertions, and deletions in each replicate population. Further, we also analyzed the genome for copy number variation. We developed a model to estimate the frequency of copy number variants in a population based on the coverage for each gene for each sample (full details in the Supplementary Information). The final list of all such genetic variants obtained for the freeze-thaw tolerant populations, along with their function, are shown in Table V. Each variant is marked for its mutation type and the sample(s) in which it was detected.

We then checked if any variants were present across multiple samples, as this would imply that the mechano-chemical phenotype found across all replicates was a consequence of a specific genotype(s). We found no variants common across all replicates of both kFT = 1 and kFT = 3 selection regimes. It must be noted that mutations in different genes with similar or shared functions, or genes linked in cellular pathways, could lead to the same adaptive phenotype. We investigated this aspect using the curated information available on the Saccharomyces genome database (SGD) on regulatory, metabolic, and physical interactions. We sought to identify a common molecular function or pathway that could have been targeted by different variants across samples. Therefore, we looked for interactions between the different genes and checked if any of the genes are regulators of other genes in the list to understand if there is any complementarity in the genotypes found such that certain replicates have related genotypes. However, no such known cellular network could be found. We also considered changes in gene-copy numbers as a potential basis for observed phenotypes, particularly the increase in intracellular levels of trehalose. Again, we found no gene with a consistent copy-number variation across all replicates (details in the Supplementary Information). These results reveal that the convergent mechano-chemical adaptations seen in this evolution experiment are achieved through multiple non-overlapping genetic routes.

Discussion

Using experimental evolution, we have shown that Saccharomyces cerevisiae can adapt to high freeze rate perturbations in a stereotypical manner, with survival increasing nearly two orders of magnitude over 25 cycles of freeze-thaw-growth. In all our replicate lines of evolution, the basal intracellular level of trehalose, a sugar speculated to play a physical role in maintaining membrane and cytoplasmic integrity [23, 56, 57], exhibited a significant (three-fold) increase in the adapted cells. The adapted cells also show smaller sizes, higher mass densities, and reduced membrane damage. Using rheological measurements we showed that the adapted cells are stiffer and had lower particle mobility within the cytoplasm. Whole genome sequencing indicated that the increased survival is associated with multiple, distinct genotypes.

Earlier, independent studies have pointed out the role of increased trehalose to confer stress tolerance in yeast, while not connecting the increased trehalose to a broader physiological context (the trehalose was increased artificially via extracellular means) [58]. Studies have also associated cytoplasmic stiffening with transition to dormancy/quiescence in yeast [19, 59] while not looking at such physical changes in an evolutionarily adaptive sense. Further, Sleight et al used experimental evolution with bacteria to shown that microbes can indeed evolve to increase survival to freeze-thaw stress – in this case, the survival increased only 2 fold, compared to the nearly 40 fold in our case – but concluded that “the physiological mechanism for this fitness improvement remains unknown” [29]. To our knowledge, our work is the first study showing the de novo linkage between increased intracellular trehalose, reduced membrane damage, decreased cytoplasmic mobility, increased cellular stiffness and transition to a quiescence-like state.

The high levels of trehalose could play multiple roles in the increased mechanical resilience of the cell. Reduced membrane damage in freeze-thaw tolerant cells could be a result of altered lipid composition, altered cell wall composition and increased trehalose. Several studies show that increased membrane fluidity of cells was associated with higher freeze-thaw tolerance, with trehalose hypothesized to aid in membrane fluidization [29, 36, 60]. In addition, our results on the increased cellular stiffness suggest that the cell wall composition could also be altered, thereby protecting the cell wall associated membrane from undergoing damage.

A major cause of death after freeze-thaw has been suggested to be ice crystal formation [23, 26, 61, 62]. Ice crystals can damage the cell membrane, proteins in the cytoplasm and disrupt the intracellular structure and organisation, thereby hindering the revival of cellular processes upon thawing. A glassy cytoplasm with decreased mobility is a cytoplasmic rheological state that could aid in retaining cellular structure and organisation during freeze-thaw transitions. Indeed such a cytoplasmic state has been implicated in response to various environmental stresses [19, 20, 43], also to be a role of trehalose [57] and has long been associated with freeze-thaw tolerance. We found the de novo encoding of a similar cytoplasmic state with increased intracellular trehalose in the adapted freeze-thaw tolerant cells.

Trehalose has been shown to be a good glass-former, compared to other sugars like glucose and sucrose [63] and has also been found to localise in the cytosol [55]. Several other studies shown that higher intracellular trehalose levels are associated with tolerance to desiccation, heat, ethanol etc [58, 64]. Cakar et al. allowed mutagenic lines of Saccharomyces cerevisiae to adapt to various kinds of stresses and found that freeze-thaw tolerant populations were tolerant to multiple other kinds of stress such as high temperature, ethanol and oxidative stress [65]. The evolution of a mechano-chemically reinforced survival phenotype underscores the role of biophysical constraints in shaping cellular adaptation. Cytoplasmic stiffening and increased trehalose accumulation likely serve as protective mechanisms against mechanical and osmotic damage [19, 20]. This highlights how stress tolerance is not solely dictated by genetic changes but also by modulations in intracellular material properties [13]. Beyond yeast, similar adaptations have been observed in bacterial biofilms and dormant cancer cells [66–68], suggesting broader implications for microbial resilience, antibiotic tolerance, and tumor dormancy.

Interestingly, our sequencing analysis did not reveal any mutations directly related to trehalose synthesis, degradation or transport (Table V). Therefore, this increased concentration of trehalose is not of any direct genetic origin, unlike previous studies that used trehalose synthesis, degradation or transport mutants and related viability after freeze-thaw to intracellular trehalose concentration [31–33].

While freeze-thaw is lethal to most kinds of cells, several cells survive freeze-thaw on ecological and evolutionary timescales. There have been a few previous studies that used experimental evolution to obtain freeze-thaw tolerance in cells. Kwon et al. started with a L. rhamnosus strain that had a 60% survival fraction after freeze-thaw and the adapted strain was found to have a ≈1.5 fold increase in survival fraction over 150 freeze-thaw-growth cycles [60]. Sleight et al. found only a 2.5 times increase in survival fraction of E. coli over 150 freeze-thaw-growth cycles (1000 generations) [29]. Although, these studies are on very different species, they both find cls mutants that increase membrane fluidity to be associated with survival. Since these kinds of stress do not cause a drastic reduction in survival, it could lead to a more gradual phenotypic change as fitness is not affected as much – it must be noted that in our experiments, the survival of wild type Saccharomyces cerevisiae to rapid freeze-thaw is a few percent or lesser. By using freeze-thaw regimes where survival is affected significantly, we found that Saccharomyces cerevisiae adapts to have orders of magnitude higher survival in only ≈25 freeze-thaw-growth cycles (≈100 generations). Our observations of stark phenotypic changes could be because of the strong selection regime in which we operate.

The various cellular attributes that we have discovered in our study correlate with those of quiescent cells. The molecular mechanisms of quiescence entry and exit and the characteristics of the quiescent state are only recently being understood [41, 47] — specifically it has been uncovered that trehalose promotes efficient entry into and exit from quiescence. Our simulations on trehalose-mediated selection for quiescence suggest that an entry to a quiescence-like state, as well as the ease of re-entering the growth cycle (through lag) facilitates the adaptation dynamics we see in the experiments.

Several studies have found that quiescent cells are more stress resilient [69], although the underlying mechanisms underlying have not yet been understood. Specifically, the genetic basis for quiescence is not fully explored. While certain genes involved in the stress response leading to quiescence have been studied through enrichment of quiescent cells in the population [70], these genetic mutations are not present among the adapted populations in our experiments. The emergence of a quiescence-like state in our experimental evolution study, suggests that increased mechanical resilience of a quiescent cell, rather than any specific genetic basis, could be responsible for stress resilience.

By bridging experimental evolution, biophysical characterization, and predictive modeling, our work provides a framework for investigating how cells can rapidly acquire stress tolerance through state transitions rather than relying solely on genetic mutations. Future research should explore the molecular regulation of these adaptive states, including potential epigenetic contributions and metabolic shifts that facilitate production of high trehalose and quiescence entry and maintenance under extreme conditions [71].